Comparing and combining network-guided biomarker discovery to discover genetic susceptibility mechanisms to breast cancer on the GENESIS study

### 2020-10-12 manuscript/plos/review_1.tex
fixed typos

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Systems biology provides a comprehensive approach to uncovering the genetics of complex diseases and building hypotheses. Indeed, it allows to jointly consider the statistical association between genetic variation and a phenotype, and the biological context of each gene, represented as a network. In this work, we study six network methods which identify subnetworks with high association scores to a phenotype. Specifically, we examine their utility to discover new breast cancer susceptibility genes by interrogating a genome-wide association study (GWAS) focused on French women with a family history of breast cancer and tested negative for pathogenic variants in \emph{BRCA1} and \emph{BRCA2}. We perform an in-depth benchmarking of the methods with regards to the size of the solution, their predictive power, and the stability and the runtime of the methods. Network methods include prior knowledge in the analysis: they might tolerate selecting a gene with a high P-value, as long as it connects other two, low P-value genes. Thanks to this, most network methods give more compelling results than standard SNP- and gene-level analyses, recovering causal subnetworks tightly related to cancer susceptibility. For instance, at least two network methods selected 14 neighbors of \emph{COPS5}, a gene related to multiple hallmarks of cancer. Importantly, we find significant overlaps between the genes in five of the solution networks and the genes significantly associated in the largest GWAS on susceptibility to breast cancer. Yet, network methods are notably unstable: the average Pearson correlation across all the methods between the genes selected on different subsamples of the data is 0.45, versus 0.57 simply using Bonferroni correction. To account for that, we produce a stable consensus solution, formed by the most consistently selected genes. The stable consensus is composed of 68 genes, enriched in known breast cancer susceptibility genes (\emph{BLM}, \emph{CASP8}, \emph{CASP10}, \emph{DNAJC1}, \emph{FGFR2}, \emph{MRPS30}, and \emph{SLC4A7}, Fisher's exact test P\=/value~=~$3 \times 10^{-4}$) and occupying more central positions in the network than the average gene. The network seems organized around \emph{CUL3}, encoding a ubiquitin ligase related protein that regulates the protein levels of several genes involved in cancer progression. In conclusion, this article shows the pertinence of network-based analyses to tackle known issues with GWAS, namely a lack of statistical power and interpretable solutions. Project-agnostic implementations of each of the network methods are available at \url{https://github.com/hclimente/gwas-tools} to facilitate their application to other GWAS datasets.

@@

In human health, genome-wide association studies (GWAS) aim at quantifying how single-nucleotide polymorphisms (SNPs) predispose to complex diseases, like diabetes or some forms of cancer \cite{bush_chapter_2012}. To that end, in a typical GWAS thousands of unrelated samples are genotyped: the cases, suffering the disease of interest, and the controls from the general population. Then, a per-SNP statistical association test is conducted (e.g. based on logistic regression). Those SNPs with a P\=/value lower than a conservative Bonferroni threshold are candidates to further studies in an independent cohort. Once the risk SNPs have been discovered, they can be used for risk assessment, and to deepen our understanding of the disease.

GWAS have successfully identified thousands of variants underlying many common diseases \cite{buniello_nhgri-ebi_2019}. However, this experimental setting also presents intrinsic challenges. Some of them stem from the high dimensionality of the problem, as every GWAS to date studies more variants than samples are genotyped. This limits the statistical power of the experiment, as only variants with larger effects can be detected \cite{visscher_10_2017}. This is particularly problematic since the prevailing view is that most genetic architectures involve many variants with small effects \cite{visscher_10_2017}. Additionally, to avoid false positives, a conservative multiple test correction is applied, typically the previously mentioned Bonferroni correction. However, Bonferroni correction is overly conservative when the statistical tests are correlated, as it is the case in GWAS \cite{wang_statistical_2018}. Another open issue is the interpretation of the results, as the functional consequences of most common variants are not well understood. On top of that, recent large-sampled studies suggest that numerous loci spread all along the genome contribute to a degree to any complex trait, in accordance with the infinitesimal model \cite{barton_infinitesimal_2017}. The recently proposed omnigenic model \cite{boyle_expanded_2017} offers an explanation: genes are very functionally inter-related and influence each other's behavior, which allows alterations in most genes to impact the subset of ``core'' genes directly involved in the disease's mechanism. Hence, a comprehensive statistical framework which includes the structure of biological data might address the aforementioned issues.

@@

Network-based discovery methods exploit the differential properties described above to identify disease genes on GWAS data \cite{azencott_network-guided_2016}. In essence, each gene has a score of association with the disease, given by the experiment, and functionally biological relationships, given by a network built on prior knowledge. Then, the problem becomes finding a functionally-related set of highly-scoring genes. Different solutions have been proposed to this problem, often stemming from divergent mathematical frameworks and considerations of what the optimal solution looks like. Some methods strongly constrain the problem to certain kinds of subnetworks. Such is the extreme case of LEAN \cite{gwinner_network-based_2016}, which focuses on ``star'' subnetworks, i.e. instances were both a gene and its direct interactors are associated with the disease. Other algorithms, like dmGWAS \cite{jia_dmgwas:_2011} and heinz \cite{dittrich_identifying_2008}, do not impose such strong constraints, and search for subnetworks interconnecting genes with high association scores. However, they differ in their tolerance to the inclusion of low-scoring nodes, and the topology of the solution. Lastly, other methods also consider the topology of the network, favoring groups of nodes that are not only high-scoring, but also densely interconnected; such is the case of HotNet2 \cite{leiserson_pan-cancer_2015}, SConES \cite{azencott_efficient_2013}, and SigMod \cite{liu_sigmod:_2017}.

In this work, we analyze the effectiveness of these six network methods on GWAS data. They do so by using different interpretations of the omnigenic model, providing a representative view of the field. We worked on the GENESIS dataset \cite{sinilnikova_genesis:_2016}, a study on familial breast cancer conducted in the French population. After a classical GWAS approach, we use these network methods to recover additional breast cancer susceptibility genes. Lastly, we carry out a comparison of the solutions obtained by the different methods, and aggregate them to obtain a consensus solution of predisposition to familial breast cancer.

@@

We applied six network methods to the GENESIS dataset (Section~\ref{methods:methods}). As none of the networks examined by LEAN was significant (Benjamini-Hochberg [BH] correction adjusted P\=/value $<$ 0.05), we obtained six solutions (Fig~\ref{fig:solution_overview}): one for each of the remaining four gene-based methods, one for SConES GI (which works at the SNP level), and the consensus. These solutions differ in many aspects, making it hard to draw joint conclusions. For starters, the overlap between the genes featured in each solution is quite small (Fig~\ref{fig:solution_overview}A). Despite that, the methods tend to agree on the genes with the strongest signal: genes selected by more methods tended to have lower P-value of association (Fig~\ref{fig:solution_overview}B). Another prominent difference is the solution size: the largest solution, produced by HotNet2, contains 440 genes, while heinz's contained only 4 genes. While SConES GI did not recover any protein coding gene, by dealing with SNP networks it retrieved four subnetworks in intergenic regions, and another one overlapping an RNA gene (\emph{RNU6-420P}). Their topologies also differ (Fig~\ref{fig:solution_overview}C), as measured by the median centrality  and the number of connected components (Table~\ref{tab:gene_solutions}). Only two methods have more than one connected component: SConES, as described above, and HotNet2. HotNet2 produced 135 subnetworks, 115 of which have less than five genes. The second largest subnetwork (13 nodes) contains the two breast cancer susceptibility genes \emph{CASP8} and \emph{BLM}. Lastly, a pathway enrichment analysis (Section \ref{methods:pathway_enrichment}) also showed similarities and differences between the solutions of the different methods. It linked different parts of SigMod's solution network to four processes (\nameref{stab:sigmod_pwy}): protein translation (including mitochondrial), mRNA splicing, protein misfolding, and keratinization (BH adjusted P\=/values~$<$~0.03). Interestingly, the dmGWAS solution (\nameref{stab:dmgwas_pwy}) is also related to protein misfolding (\emph{attenuation phase}, BH adjusted P\=/value~=~0.01). But, additionally, it includes submodules of proteins related to mitosis, DNA damage, and regulation of TP53 (BH adjusted P\=/values~$<$~0.05), which match previously known mechanisms of breast cancer susceptibility \cite{nielsen_hereditary_2016}. As with SigMod, the genes in HotNet2's solution (\nameref{stab:hotnet2_pwy}) are involved in mitochondrial translation (BH adjusted P\=/value~=~1.87 \texttimes{} 10\textsuperscript{-4}), but also in glycogen metabolism and transcription of nuclear receptors (BH adjusted P\=/value~$<$~0.04).

@@

Despite their differences, there are additional common themes. All obtained solutions have lower association P\=/values than the whole PPIN (median VEGAS2 P\=/value $\ll 0.46$, Table~\ref{tab:gene_solutions}), despite containing genes with higher P\=/values as well (Fig~\ref{fig:solution_overview}D). This exemplifies the trade-off between controlling for type I error and biological relevance. However, there are nuances between solutions in this regard: heinz strongly favored genes with lower P\=/values, while dmGWAS was less conservative (median VEGAS2 P\=/values 0.0012 and 0.19, respectively); SConES tended to select whole LD-blocks; and HotNet2 and SigMod were less likely to select low scoring genes. Additionally, the solutions presented other desirable properties. First, five of them were enriched in known breast cancer susceptibility genes (consensus, dmGWAS, heinz, HotNet2, and SigMod, Fisher's exact test one-sided P\=/value$<$0.03). Second, the genes in four solutions displayed on average a higher betweenness centrality than the rest of the genes, a difference that is significant in four solutions (consensus, dmGWAS, HotNet2, and SigMod, Wilcoxon rank-sum test P\=/value$<$1.4 \texttimes{} 10\textsuperscript{-21}). This agrees with the notion that disease genes are more central than other non-essential genes \cite{pinero_uncovering_2016}, an observation that holds in breast cancer (one-tailed Wilcoxon rank-sum test P\=/value~=~2.64 \texttimes{} 10\textsuperscript{-5} when comparing the betweenness of known susceptibility genes versus the rest). Interestingly, SConES selected SNPs that are also more central than the average SNP (\nameref{stab:snp_solutions}), suggesting that causal SNPs are also more central than non-associated SNPs.

@@

Topologically the consensus consists of a connected component composed of 49 genes, and multiple smaller subnetworks. Among the latter, 19 genes are in subnetworks containing a single gene or two connected nodes, implying that they do not have a consistently altered neighborhood, but are strongly associated themselves and hence picked by at least two methods. In the former we find some genes that are highly central in the PPIN, a property which is weakly anti-correlated with the P\=/value of association to the disease (Pearson correlation coefficient~=~-0.26, \nameref{sfig:consensus_stats}). This suggests that these genes were selected because they were on the shortest path between another two highly associated genes. In view of this, we hypothesize that highly central genes might contribute to the heritability through alterations of their neighborhood, consistently with the omnigenic model of disease \cite{boyle_expanded_2017}. For instance, the most central node in the consensus solution is COPS5, a component of the COP9 signalosome which regulates multiple signaling pathways. \emph{COPS5} is related to multiple hallmarks of cancer and is overexpressed in multiple tumors, including breast and ovarian cancer \cite{liu_jab1_cops5_2018}. Despite its lack of association in GENESIS or BCAC (VEGAS2 P\=/value of 0.22 and 0.14 respectively), its neighbors in the consensus solution have consistently low P\=/values (median VEGAS2 P\=/value~=~0.006).

@@

We compared the results of different network methods to the European sample of the Breast Cancer Association Consortium (BCAC) \cite{Michailidou2017}, the largest GWAS to date on breast cancer (Section~\ref{methods:bcac}). Although BCAC case-control studies do not necessarily target cases with a family history of breast cancer, this comparison is pertinent since we expect a shared genetic architecture at the gene level, at which most network methods operate. This shared genetic architecture, together with BCAC's scale (90 times more samples than GENESIS) provides a reasonable counterfactual of what we would expect if GENESIS had a larger sample size. We computed a gene association score on BCAC (Section~\ref{methods:bcac}). The solutions provided by the different  network methods overlap significantly with BCAC findings (Fisher's exact test P\=/value \textless~0.019). The gene-based methods achieve comparable precision (2\%-25\%) and recall (1.3-12.1\%) at recovering BCAC-significant genes (\nameref{sfig:additional-benchmarks}A). interestingly, while scones gi at the snp-level achieves a similar recall (8.6\%), it shows a much higher precision (47.3\%).

@@

We compared the six network methods in a 5-fold subsampling setting (Section \ref{methods:benchmark}). Specifically, we measured five properties (Fig~\ref{fig:benchmark}): size of the solution; sensitivity and specificity of an L1-penalized logistic regression classifier on the selected SNPs; stability; and computational runtime. The solution size varies greatly between the different methods (Fig~\ref{fig:benchmark}A). Heinz produced the smallest solutions, with an average of 182 selected SNPs (Section~\ref{methods:snp2gene}). The largest solutions came from SConES GI (6\,256.6 SNPs), and dmGWAS (4\,255.0 SNPs). LEAN did not produce any solution in any of the subsamples. To determine whether the selected snps could be used for patient classification, we computed the performance of the classifier on the \emph{test dataset} (\nameref{sfig:additional-benchmarks}B). The different classifiers displayed similarly poor sensitivities and specificities, all in the 0.52 -- 0.56 range. Interestingly, the classifier trained on all the SNPs had a similar performance, despite being the only method aiming only at minimizing prediction error. It should be considered that, although these performances are low, we do not expect to separate cases from controls well using exclusively genetic data \cite{deloscamposComplexTraitPredictionEra2018}.

@@

Another desirable quality of an algorithm is the stability of the solution with regards to small changes in the input (Section~\ref{methods:algorithm_comparison}). Heinz was high stabile in our benchmark, while the other methods displayed similarly low stabilities (Fig~\ref{fig:benchmark}C).

In terms of computational runtime, the fastest method was heinz (Fig~\ref{fig:benchmark}D), which returned a solution in a few seconds. HotNet2 was the slowest (3 days and 14 hours on average). Including the time required to compute the gene scores, however, slows down considerably gene-based methods; on this benchmark, that step took on average 1 day and 9.33 hours. Considering that, it took 5 days on average for HotNet2 to produce a result.

@@

As shown above, and despite their similarities, the network methods produced remarkably different solutions. This is due to the fact that each of them models the optimal solution differently. Importantly, understanding which assumptions they make allows to understand the dataset more deeply. For instance, the fact that LEAN did not return any gene implies that there is no gene such that both itself and its environment are on average strongly associated with the disease.

@@

Another peculiarity of network methods is their relationship to degree centrality. We studied randomly rewirings of the PPIN while preserving node centrality (Section~\ref{methods:rewiring}). In this setting, network methods favored central genes, as they often connect high scoring nodes (Fig~\ref{fig:issues}B). This is despite the fact that highly central genes often had no association to breast cancer susceptibility (Fig~\ref{fig:issues}C). This was specially the case of SigMod, which selected three highly central, unassociated genes: \emph{COPS5}, \emph{CUL3} and \emph{FN1}. As we showed in Section \ref{results:consensus}, and will show in \ref{results:stable-consensus}, there is evidence in the literature of the contribution of the first two to breast cancer susceptibility. With regards to \emph{FN1}, it encodes a fibronectin, a protein of the extracellular matrix involved in cell adhesion and migration. Overexpression of \emph{FN1} has been observed in breast cancer \cite{Ioachim2002}, and it is negatively correlated with poor prognosis in other cancer types \cite{Yi2016,Sponziello2016}.

By virtue of using a SNP subnetwork, SConES analyzes each SNP in their functional context. It therefore can select SNPs in genes without any associated interactor, as well as SNPs in non-coding regions or in non-interacting genes. We compared the solution of SConES in the GI network (using PPIN information), to the one using only positional information (GS network), or positional and gene annotations (GM network). Importantly, SConES produces similar results on the GS and GM networks (\nameref{sfig:pearson_methods}). While the solutions on those two greatly overlap with SConES GI's, they contain additional gene-coding segments (Fig~\ref{fig:issues}C). In fact the formers' solutions are chromosome regions related to breast cancer, like 3p24 (\emph{SLC4A7}/\emph{NEK10} \cite{ahmed_newly_2009}), 5p12 (\emph{FGF10}, \emph{MRPS30} \cite{quigley_5p12_2014}), 10q26 (\emph{FGFR2}), and 16q12 (\emph{TOX3}). On top of those SConES GS selects region 8q24 (\emph{POU5F1B} \cite{breyer_expressed_2014}).

\subsection{Adjusting for instability preserves global network properties}
\label{results:stable-consensus}

@@

Despite these similarities, the consensus and the stable consensus solutions include different genes. In this case, the most central gene is \emph{CUL3}, which is absent from the previous consensus solution and has a low association score in both GENESIS and BCAC (P\=/values of 0.04 and 0.26, respectively). This gene is a component of Cullin-RING ubiquitin ligases. Encouragingly, it impacts the protein levels of multiple genes relevant for cancer progression \cite{Chen2016}, and its overexpression was also linked to increased sensitivity to carcinogens \cite{Loignon2009}.

@@

In recent years, the ability of GWAS to unravel the mechanisms leading to complex diseases has been called into question \cite{boyle_expanded_2017}. On the one hand, the omnigenic model proposes that gene functions are interwoven with each other in a dense co-function network. The practical consequence is that larger and larger GWAS will lead to the discovery of an uninformative wide-spread pleiotropy. On the other hand, discovery in GWAS is hindered by a conservative statistical framework. Network methods tackle these two issues by using both the association score and an interaction network to take into consideration the biological context of each of the genes and SNPs. Based on what could be considered diverse interpretations of the omnigenic model, several methods for network-guided discovery have been proposed in recent years. In this article we evaluated the relevance of six of them by examining GENESIS, a GWAS dataset on familial breast cancer.

>>>>

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Systems biology provides a comprehensive approach to uncovering the genetics of complex diseases and building hypotheses. Indeed, it allows to jointly consider the statistical association between genetic variation and a phenotype, and the biological context of each gene, represented as a network. In this work, we study six network methods which identify subnetworks with high scores of association with a phenotype. More specifically, we examine their use to discover new breast cancer susceptibility genes by interrogating a genome-wide association study (GWAS) data set focused on French women with a family history of breast cancer and tested negative for pathogenic variants in \emph{BRCA1} and \emph{BRCA2}. We perform an in-depth benchmarking of these six methods with respect to the size of the solution, its predictive power, and the stability and the runtime of the methods. Network methods include prior knowledge in the analysis: they might tolerate selecting a gene with a high P-value, as long as it connects other two, low P-value genes. This alows them to give more compelling results than standard SNP- and gene-level analyses, recovering causal subnetworks tightly related to cancer susceptibility. For instance, at least two network methods selected 14 neighbors of \emph{COPS5}, a gene related to multiple hallmarks of cancer. Importantly, we find significant overlaps between the genes in five of the solution networks and the genes significantly associated in the largest GWAS on susceptibility to breast cancer. Yet, network methods are notably unstable: the average Pearson correlation across all the methods between the genes selected on different subsamples of the data is 0.45, versus 0.57 simply using Bonferroni correction. To compensate this instability, we produce a stable consensus solution, formed by the most consistently selected genes. This stable consensus is composed of 68 genes, enriched in known breast cancer susceptibility genes (\emph{BLM}, \emph{CASP8}, \emph{CASP10}, \emph{DNAJC1}, \emph{FGFR2}, \emph{MRPS30}, and \emph{SLC4A7}, Fisher's exact test P\=/value~=~$3 \times 10^{-4}$) and occupying more central positions in the network than the average gene. The network seems organized around \emph{CUL3}, which encodes a ubiquitin ligase related protein that regulates the protein levels of several genes involved in cancer progression. In conclusion, this article shows the pertinence of network-based analyses to address known limitations of GWAS, namely their lack of statistical power and the difficulty of their interpretation. Project-agnostic implementations of each of the network methods are available at \url{https://github.com/hclimente/gwas-tools} to facilitate their application to other GWAS datasets.

@@

In human health, genome-wide association studies (GWAS) aim at quantifying how single-nucleotide polymorphisms (SNPs) predispose to complex diseases, like diabetes or some forms of cancer \cite{bush_chapter_2012}. To that end, in a typical GWAS thousands of unrelated samples are genotyped: the cases, suffering from the disease of interest, and the controls, taken from the general population. Then, a statistical test of association (e.g. based on logistic regression) is conducted between each individual SNP and the phenotype. Those SNPs with a P\=/value lower than a conservative Bonferroni threshold are candidates to further studies in an independent cohort. Once the risk SNPs have been discovered, they can be used for risk assessment, and to deepen our understanding of the disease.

GWAS have successfully identified thousands of variants underlying many common diseases \cite{buniello_nhgri-ebi_2019}. However, this experimental setting also presents intrinsic challenges. Some of them stem from the high dimensionality of the problem, as every GWAS to date studies more variants than samples are genotyped. This limits the statistical power of the experiment, as only variants with larger effects can be detected \cite{visscher_10_2017}. This is particularly problematic since the prevailing view is that most genetic architectures involve many variants with small effects \cite{visscher_10_2017}. Additionally, to avoid false positives, a conservative multiple test correction is applied, typically the previously mentioned Bonferroni correction. However, Bonferroni correction is overly conservative when the statistical tests are correlated, as it is the case in GWAS \cite{wang_statistical_2018}. Another open issue is the interpretation of the results, as the functional consequences of most common variants are not well understood. On top of that, recent large-sampled studies suggest that numerous loci spread all along the genome contribute to a degree to any complex trait, in accordance with the infinitesimal model \cite{barton_infinitesimal_2017}. The recently proposed omnigenic model \cite{boyle_expanded_2017} offers an explanation: genes are strongly functionally inter-related and influence each other's behavior, which allows alterations in most genes to impact the subset of ``core'' genes directly involved in the disease's mechanism. Hence, a comprehensive statistical framework which includes the structure of biological data might help alleviate the aforementioned issues.

@@

Network-based discovery methods exploit the differential properties described above to identify disease genes on GWAS data \cite{azencott_network-guided_2016}. In essence, each gene is assigned a score of association with the disease, computed from the GWAS data, and biological relationships, given by a network built on prior knowledge. Then, the problem becomes finding a functionally-related set of highly-scoring genes. Multiple solutions have been proposed to this problem, often stemming from different mathematical frameworks and considerations of what the optimal solution looks like. For example, some methods restrict the problem to specific types of subnetworks. Such is the case of LEAN \cite{gwinner_network-based_2016}, which focuses on ``star'' subnetworks, i.e. instances were both a gene and its direct interactors are associated with the disease. Other algorithms, like dmGWAS \cite{jia_dmgwas:_2011} and heinz \cite{dittrich_identifying_2008}, do not impose such strong constraints, and search for subnetworks interconnecting genes with high association scores. However, they differ in their tolerance to the inclusion of low-scoring nodes, and the topology of the solution. Lastly, other methods also consider the topology of the network, favoring groups of nodes that are not only high-scoring, but also densely interconnected; such is the case of HotNet2 \cite{leiserson_pan-cancer_2015}, SConES \cite{azencott_efficient_2013}, and SigMod \cite{liu_sigmod:_2017}.

In this work, we analyze the application of these six network methods on GWAS data. They do so by using different interpretations of the omnigenic model, providing a representative view of the field. We worked on the GENESIS dataset \cite{sinilnikova_genesis:_2016}, a study on familial breast cancer conducted in the French population. After a classical GWAS approach, we use these network methods to recover additional breast cancer susceptibility genes. Lastly, we carry out a comparison of the solutions obtained by the different methods, and aggregate them to obtain a consensus solution of predisposition to familial breast cancer.

@@

We applied six network methods to the GENESIS dataset (Section~\ref{methods:methods}). As none of the networks examined by LEAN was significant (Benjamini-Hochberg [BH] correction adjusted P\=/value $<$ 0.05), we obtained six solutions (Fig~\ref{fig:solution_overview}): one for each of the remaining four gene-based methods, one for SConES GI (which works at the SNP level), and the consensus.

These solutions differ in many aspects, making it hard to draw joint conclusions. For starters, the overlap between the genes featured in each solution is quite small (Fig~\ref{fig:solution_overview}A). However, the methods tend to agree on the genes with the strongest signal: genes selected by more methods tended to have lower P-value of association (Fig~\ref{fig:solution_overview}B).

Another prominent difference is the solution size: the largest solution, produced by HotNet2, contains 440 genes, while heinz's contains only 4 genes. While SConES GI did not recover any protein coding gene, working with SNP networks rather than gene networks allowed it to retriev four subnetworks in intergenic regions, and another one overlapping an RNA gene (\emph{RNU6-420P}).

The topologies of the six solutions differ as well (Fig~\ref{fig:solution_overview}C), as measured by the median centrality  and the number of connected components (Table~\ref{tab:gene_solutions}). Only two methods yield more than one connected component: SConES, as described above, and HotNet2. HotNet2 produced 135 subnetworks, 115 of which have fewer than five genes. The second largest subnetwork (13 nodes) contains the two breast cancer susceptibility genes \emph{CASP8} and \emph{BLM}.

Lastly, a pathway enrichment analysis (Section \ref{methods:pathway_enrichment}) also showed similarities and differences between the solutions of the different methods. It linked different parts of SigMod's solution to four processes (\nameref{stab:sigmod_pwy}): protein translation (including mitochondrial), mRNA splicing, protein misfolding, and keratinization (BH adjusted P\=/values~$<$~0.03). Interestingly, the dmGWAS solution (\nameref{stab:dmgwas_pwy}) is also related to protein misfolding (\emph{attenuation phase}, BH adjusted P\=/value~=~0.01). However, it additionally includes submodules of proteins related to mitosis, DNA damage, and regulation of TP53 (BH adjusted P\=/values~$<$~0.05), which match previously known mechanisms of breast cancer susceptibility \cite{nielsen_hereditary_2016}. As with SigMod, the genes in HotNet2's solution (\nameref{stab:hotnet2_pwy}) are involved in mitochondrial translation (BH adjusted P\=/value~=~1.87 \texttimes{} 10\textsuperscript{-4}), but also in glycogen metabolism and transcription of nuclear receptors (BH adjusted P\=/value~$<$~0.04).

@@

Despite their differences, there are additional common themes. All obtained solutions have lower association P\=/values than the whole PPIN (median VEGAS2 P\=/value $\ll 0.46$, Table~\ref{tab:gene_solutions}), despite containing genes with higher P\=/values as well (Fig~\ref{fig:solution_overview}D). This illustrates the trade-off between controlling for type I error and biological relevance. However, there are nuances between solutions in this regard: heinz strongly favors genes with lower P\=/values, while dmGWAS is less conservative (median VEGAS2 P\=/values 0.0012 and 0.19, respectively); SConES tends to select whole LD-blocks; and HotNet2 and SigMod are less likely to select low scoring genes.

Additionally, the solutions presented other desirable properties. First, five of them were enriched in known breast cancer susceptibility genes (consensus, dmGWAS, heinz, HotNet2, and SigMod, Fisher's exact test one-sided P\=/value$<$0.03). Second, the genes in four solutions displayed on average a higher betweenness centrality than the rest of the genes, a difference that is significant in four solutions (consensus, dmGWAS, HotNet2, and SigMod, Wilcoxon rank-sum test P\=/value$<$1.4 \texttimes{} 10\textsuperscript{-21}). This agrees with the notion that disease genes are more central than other non-essential genes \cite{pinero_uncovering_2016}, an observation that holds in breast cancer (one-tailed Wilcoxon rank-sum test P\=/value~=~2.64 \texttimes{} 10\textsuperscript{-5} when comparing the betweenness of known susceptibility genes versus the rest). Interestingly, SConES selected SNPs that are also more central than the average SNP (\nameref{stab:snp_solutions}), suggesting that causal SNPs are also more central than non-associated SNPs.

@@

Topologically, the consensus consists of a connected component composed of 49 genes, and multiple smaller subnetworks. Among the latter, 19 genes are in subnetworks containing a single gene or two connected nodes, implying that they do not have a consistently altered neighborhood, but are strongly associated themselves and hence picked by at least two methods. The large connected component contains genes that are highly central in the PPIN, a property which is weakly anti-correlated with the P\=/value of association to the disease (Pearson correlation coefficient~=~-0.26, \nameref{sfig:consensus_stats}). This suggests that these genes were selected because they were on the shortest path between another two highly associated genes. In view of this, we hypothesize that highly central genes might contribute to the heritability through alterations of their neighborhood, consistently with the omnigenic model of disease \cite{boyle_expanded_2017}. For instance, the most central node in the consensus solution is COPS5, a component of the COP9 signalosome which regulates multiple signaling pathways. \emph{COPS5} is related to multiple hallmarks of cancer and is overexpressed in multiple tumors, including breast and ovarian cancer \cite{liu_jab1_cops5_2018}. Despite its lack of association in GENESIS or BCAC (VEGAS2 P\=/value of 0.22 and 0.14 respectively), its neighbors in the consensus solution have consistently low P\=/values (median VEGAS2 P\=/value~=~0.006).

@@

We compared the results obtained with different network methods to the European sample of the Breast Cancer Association Consortium (BCAC) \cite{Michailidou2017}, the largest GWAS to date on breast cancer (Section~\ref{methods:bcac}). Although BCAC case-control studies do not necessarily target cases with a familial history of breast cancer, this comparison is pertinent since we expect a shared genetic architecture at the gene level, at which most network methods operate. This shared genetic architecture, together with BCAC's scale (90 times more samples than GENESIS) provides a reasonable counterfactual of what we would expect if GENESIS had a larger sample size. We computed a gene association score on BCAC (Section~\ref{methods:bcac}). The solutions provided by the different  network methods overlap significantly with BCAC findings (Fisher's exact test P\=/value \textless~0.019). The gene-based methods achieve comparable precision (2\%-25\%) and recall (1.3-12.1\%) at recovering BCAC-significant genes (\nameref{sfig:additional-benchmarks}A). Interestingly, while SConES GI, at the SNP-level, achieves a similar recall (8.6\%), it shows a much higher precision (47.3\%).

@@

We compared the six network methods in a 5-fold subsampling setting (Section \ref{methods:benchmark}). This allowed us to measure five properties (Fig~\ref{fig:benchmark}): size of the solution; sensitivity and specificity of an L1-penalized logistic regression classifier on the selected SNPs; stability; and computational runtime. The solution size varies greatly between the different methods (Fig~\ref{fig:benchmark}A). Heinz produced the smallest solutions, with an average of 182 selected SNPs (Section~\ref{methods:snp2gene}). The largest solutions came from SConES GI (6\,256.6 SNPs), and dmGWAS (4\,255.0 SNPs). LEAN did not produce any solution in any of the subsamples. To determine whether the selected SNPs could be used for patient classification, we computed the performance of the classifier on the \emph{test dataset} (\nameref{sfig:additional-benchmarks}B). The different classifiers displayed similarly poor sensitivities and specificities, all in the 0.52 -- 0.56 range. Interestingly, the classifier trained on all the SNPs had a similar performance, despite being the only method aiming only at minimizing prediction error. Of course, although these performances are low, we do not expect to separate cases from controls well using exclusively genetic data \cite{deloscamposComplexTraitPredictionEra2018}.

@@

Another desirable quality of a selection algorithm is the stability of the solution with respect to small changes in the input (Section~\ref{methods:algorithm_comparison}). Heinz was highly stable in our benchmark, while the other methods displayed similarly low stabilities (Fig~\ref{fig:benchmark}C).

In terms of computational runtime, the fastest method was heinz (Fig~\ref{fig:benchmark}D), which returned a solution in a few seconds. HotNet2 was the slowest (3 days and 14 hours on average). Including the time required to compute the gene scores, however, slows down considerably gene-based methods; on this benchmark, that step took on average 1 day and 9.33 hours. Including this first step, it therefore took 5 days on average for HotNet2 to produce a result.

@@

As shown above, and despite their similarities, the network methods produced remarkably different solutions. This is due to the fact that each of them models the problem differently. Importantly, understanding which assumptions they make allows to understand the results more in depth. For instance, the fact that LEAN did not return any gene implies that there is no gene such that both itself and its environment are on average strongly associated with the disease.

@@

Another peculiarity of network methods is their relationship to degree centrality. We studied random rewirings of the PPIN while preserving node centrality (Section~\ref{methods:rewiring}). In this setting, network methods favored central genes, as they often connect high scoring nodes (Fig~\ref{fig:issues}B). This is despite the fact that highly central genes often had no association to breast cancer susceptibility (Fig~\ref{fig:issues}C). This was especially the case of SigMod, which selected three highly central, unassociated genes: \emph{COPS5}, \emph{CUL3} and \emph{FN1}. As we showed in Section \ref{results:consensus}, and will show in \ref{results:stable-consensus}, there is evidence in the literature of the contribution of the first two to breast cancer susceptibility. With regards to \emph{FN1}, it encodes a fibronectin, a protein of the extracellular matrix involved in cell adhesion and migration. Overexpression of \emph{FN1} has been observed in breast cancer \cite{Ioachim2002}, and it is negatively correlated with poor prognosis in other cancer types \cite{Yi2016,Sponziello2016}.

By virtue of using a SNP subnetwork, SConES analyzes each SNP in their functional context. It can therefore select SNPs located in genes not included in the PPIN, as well as SNPs in non-coding regions or in non-interacting genes. We compared the solution of SConES in the GI network (using PPIN information), to the one using only positional information (GS network), or positional and gene annotations (GM network). Importantly, SConES produces similar results on the GS and GM networks (\nameref{sfig:pearson_methods}). While the solutions on those two greatly overlap with SConES GI's, they contain additional gene-coding segments (Fig~\ref{fig:issues}C). In fact the formers' solutions are chromosome regions related to breast cancer, like 3p24 (\emph{SLC4A7}/\emph{NEK10} \cite{ahmed_newly_2009}), 5p12 (\emph{FGF10}, \emph{MRPS30} \cite{quigley_5p12_2014}), 10q26 (\emph{FGFR2}), and 16q12 (\emph{TOX3}). On top of those SConES GS selects region 8q24 (\emph{POU5F1B} \cite{breyer_expressed_2014}).

\subsection{Building a stable consensus network preserves global network properties}
\label{results:stable-consensus}

@@

Despite these similarities, the consensus and the stable consensus solutions include different genes. In the stable consensus network, the most central gene is \emph{CUL3}, which is absent from the previous consensus solution and has a low association score in both GENESIS and BCAC (P\=/values of 0.04 and 0.26, respectively). This gene is a component of Cullin-RING ubiquitin ligases. Encouragingly, it impacts the protein levels of multiple genes relevant for cancer progression \cite{Chen2016}, and its overexpression was also linked to increased sensitivity to carcinogens \cite{Loignon2009}.

@@

In recent years, the ability of GWAS to unravel the mechanisms leading to complex diseases has been called into question \cite{boyle_expanded_2017}. First, the omnigenic model proposes that gene functions are interwoven with each other in a dense co-function network. The practical consequence is that larger and larger GWAS will lead to the discovery of an uninformative wide-spread pleiotropy. Second, discovery in GWAS is hindered by a conservative statistical framework. Network methods elegantly address these two issues by using both association scores and an interaction network to take into consideration the biological context of each of the genes and SNPs. Based on what could be considered diverse interpretations of the omnigenic model, several methods for network-guided discovery have been proposed in recent years. In this article we evaluated the relevance of six of these methods by applying them to the study of GENESIS, a GWAS dataset on familial breast cancer.
